The role of advection versus diffusion in intraventricular transport

### 2025-02-04 paper/main.tex
redistribute/remove 'Physical parameters' section

\section*{Methods}
\subsection*{Image-based computational geometries of zebrafish brain ventricles}
To model flow and transport in zebrafish larval brain ventricles, we generated a 3D representation of the brain ventricular walls using confocal imaging data of a zebrafish embryo
injected intraventricularly with a 70 kDa dye at 2 dpf~\cite{Olstad2019CiliaryDevelopment}. 
From the imaging data, we constructed a computational mesh of the ventricles (\Cref{fig:figure1}d).
The interior volume was meshed with fTetWild~\cite{Hu2020FastWild}. 
The \emph{standard} mesh consisted of 132 134 tetrahedral cells and had a maximal (minimal) edge length of 14.9 $\mu$m (4.68 $\mu$m). 
At 2 dpf, the ventricular system consists mainly of 3 cavities: the telencephalic (anterior),
the di-/mesencephalic (middle) and the rhombencephalic (posterior) ventricles, which are connected by ducts. 
Parts of the ventricular surfaces were marked to be lined with
motile cilia~\cite{Olstad2019CiliaryDevelopment} (\Cref{fig:figure1}d, pink markers).
\begin{figure}%[hbt!]
    \centering
    \includegraphics[width=\linewidth]{graphics/figure1.png}
    \caption{\lyng{Experimenting with velocity colorbars, thoughts on the two options? Or have two separate as it is now?}}
    \label{fig:figure1}
\end{figure}

@@

\subsection*{Computational CSF dynamics}
The beating motion of the cilia generates CSF flow with steady rotational structures~\cite{Olstad2019CiliaryDevelopment}
within the brain ventricles (\Cref{fig:figure1}a--c).
In addition, cardiac pulsations induces pulsatile CSF flow.
We model this flow of CSF in the brain ventricles by the time-dependent,
incompressible Stokes equations which read as follows.
Find the CSF velocity $\uu = \uu(\xx, t) = (u_x(\xx, t), u_y(\xx, t), u_z(\xx, t))$
and the CSF pressure $p = p(\xx, t)$ for $\xx \in \Omega$ and time $t>0$, such that
\begin{subequations}
    \begin{align}
      \rho \pdifft{\uu} - \nabla \cdot \bsig &= \mathbf{0}
      \quad \text{in } \Omega,\\
      \nabla \cdot \uu &= 0
      \quad  \text{in } \Omega,
  \end{align}
  \label{eq:stokes_eqs}% 
\end{subequations}%
where the stress tensor is defined as $\bsig(\uu, p) = 2\mu\beps(\uu) - p\mathbf{I}$
and $\beps(\uu) = \frac{1}{2}\left(\nabla \uu + (\nabla\uu)^T\right)$ is the strain-rate tensor. 
Bold-face characters denote vectors or tensors, and $\mathbf{I}$ is the identity tensor in three dimensions.
Furthermore, $\rho=1000 \ \mathrm{kg/m^3}$ is the CSF density
and $\mu=6.97\times 10^{-4}$ Pa is the dynamic viscosity of the CSF.
We introduce the tangential traction $\bsigpar$ as the
tangential component of the traction $\hat{\bsig}=\bsig\nn$:
\begin{equation*}
    \bsigpar = P_{\nn}(\hat{\bsig}) = (\mathbf{I} - \nn\otimes\nn)\hat{\bsig},
\end{equation*}
where $\nn$ is the outer unit normal of the boundary surface $\Gamma = \partial\Omega$,
and $P_{\nn}(\rr)$ is the tangential projection of a vector $\rr$ onto $\Gamma$.
We consider three types of boundary conditions:
(i) \emph{tangential traction} (no flow normal to the boundary $\uu \cdot \nn = 0$ with a force $\btau$ applied tangentially on the boundary $\bsigpar = \btau$),
(ii) \emph{normal pressure} (a pressure $\Tilde{p}$ applied in the normal direction: $(\mu\nabla\uu - p\mathbf{I})\nn = \Tilde{p}(t)\nn$),
and (iii) \emph{free slip} (no flow normal to the boundary $\uu \cdot \nn = 0$ and no tangential forces $\bsigpar = {\bf 0}$).
Unless noted otherwise, the free-slip condition is applied on the boundary. We assume that the system starts at rest: $\uu(\xx, t=0) = \mathbf{0}$.

\subsection*{Cilia-driven flow}
On the ciliated regions of the ventricle walls $\Gc$ (\Cref{fig:figure1}d),
we impose a constant-in-time tangential traction $\bsigpar=\btau$ to represent the net forces of the cilia acting on the CSF.
We set $\btau(\xx, \rr) = \tau \lambda(\xx) P_{\nn}(\rr)$, where the function $\lambda(\xx)$
and the sign of the vector $\rr=\pm(1, 0, 1)$ varies depending on the cilia region (see~\Cref{subsec:appendixA1}).
Both $\rr$ and the value $\tau=6.5\times 10^{-4}$ Pa were chosen by heuristic
model calibration using experimental data~\cite{Olstad2019CiliaryDevelopment}.

\subsection*{Flow induced by cardiac pulsations}\label{subsec:pressure_bcs}
To drive a cardiac-induced pulsatile flow inside the brain ventricles, we set $\Tilde{p}(t)=0$ and
$\Tilde{p}(t)=-A\sin{\omega t}$ on anterior and posterior parts, respectively, of the boundary $\Gp$~(\Cref{fig:figure1}d, blue markers).
The amplitude $A=1.5\times 10^{-3}$ Pa determines the magnitude of the pulsatile flow,
and was calibrated based on experimental data~\cite{Olstad2019CiliaryDevelopment}.
The angular cardiac frequency $\omega=2\pi f=6.97$ rad/s was set based on a measured cardiac frequency of $f=2.22$ Hz.

@@

To computationally represent protein photoconversion in a region $\Omega_{\mathrm{pc}} \subset \Omega$ (corresponding to ROI 1), we prescribe a given time-dependent value to the solute concentration in this region. 
Mimicking the experimental protocol, we consider the dorsal part of the diencephalic ventricle as the subdomain $\Omega_{\mathrm{pc}}$, and set
\begin{equation}
  c(\xx, t)=\frac{\log{(1+t/a)}}{\log{(1+T/a)}}
  \qquad \xx \in \Omega_{\mathrm{pc}}, \quad \fixme{t > 0}.
\label{eq:photoconversion_curve}
\end{equation}
The value $a=65$ was chosen to fit the fluorescence intensity curve observed in the physical photoconversion experiments.

\subsection*{Estimation of diffusion coefficients via the Stokes-Einstein relation}
Motivated by an interest in molecules that are significant in neural development, we study the transport of molecules resembling Dendra2 Fluorescent Protein (Dendra2), Starmaker+Green Fluorescent Protein (STM+GFP) and Exosomes.
To estimate the diffusion coefficients of these, when and if not available in the literature, we use the Stokes--Einstein relation:
\begin{equation}
    D = \frac{k_B T}{6\pi \mu R},
    \label{eq:D_stokes_einstein}
\end{equation}
which is an expression for the diffusion coefficient of a spherical molecule suspended in a viscous fluid~\cite{Einstein1905UberTeilchen}.
In~\eqref{eq:D_stokes_einstein}, $k_B = \num{1.38e-23}$ J/K is the Boltzmann constant, $T = 310$ K is the absolute temperature, $\mu$ is the dynamic viscosity of the solvent (in this case the CSF), and $R$ is the radius of the molecule.
For the exosomes, we calculate $D$ using~\eqref{eq:D_stokes_einstein} with $R=200$ nm. This radius covers the slowest diffusing exosomes, since the smaller exosomes would have diffusion coefficients similar to the other molecules.
For STM+GFP, we estimated the diffusion coefficient by extrapolating the experimentally observed diffusion coefficient of GFP~\cite{Swaminathan1997PhotobleachingDiffusion, Potma2001ReducedCells},
assuming that if $D$ scales with $1/R$ according to~\eqref{eq:D_stokes_einstein}, $D$ would scale with the cube root of the molecule mass~\cite{Goodhill1997DiffusionGuidance}.
The resulting diffusion coefficients used in the simulations are summarized in \Cref{tab:diff_coeff_table}.

@@

\subsection*{Physical parameters}
The CSF density $\rho$ and dynamic viscosity $\mu$ were set following Bloomfield \textit{et al.}~\cite{Bloomfield1998EffectsFluid}~(\Cref{tab:parameter_table}). We considered large, medium and small-sized molecules by three different diffusion coefficients $D_1, D_2$ and $D_3$ (\Cref{tab:parameter_table}). Experimental data~\cite{Olstad2019CiliaryDevelopment} indicate \fixme{velocities} proximate to the ventricular walls of $27.4 \pm 5.4$ \textmu m/s and $2.6 \pm 0.6$ \textmu m/s in the dorsal and ventral regions of the diencephalic ventricle, respectively. In the telencephalic ventricle, the velocities were estimated at $4.7 \pm 1.3$ \textmu m/s. We calibrated $\tau$ and $A$ heuristically by comparing these experimental velocity values with the simulated, maximum velocity magnitude. In addition, particle tracking data of pulsatile motion was used in the calibration process to determine the relative contributions of the cardiac pulsations, as compared to the cilia-driven flow, to the velocity magnitude.
\mer{MER: What velocities? Net velocities? Velocity magnitude aka speed? Also, haven't you said this part about the experimental data already? Either place the details there, or all here. I suggest there.}
\mer{MER: I suggest you introduce parameter values when introduced, and move the parts relevant for cilia/cardiac flow to where you mentioned it first, and then merge this section with the Estimation of diffusion .... Then remove this section and \Cref{tab:parameter_table}.}
\begin{table}[h!]
    \centering
    \begin{tabular}{lcccc}
    \toprule
        Parameter & Symbol & Value & Unit & Reference \\
        \midrule
        Fluid density & $\rho$ & $1000$ & $\mathrm{kg/m^3}$ & \cite{Bloomfield1998EffectsFluid} \\
        Dynamic viscosity & $\mu$ & $\num{6.97e-04}$ & $\mathrm{Pa\cdot s}$ & \cite{Bloomfield1998EffectsFluid} \\
        Cardiac frequency & $f$ & 2.22 & Hz &  \cite{Olstad2019CiliaryDevelopment}\\
        Angular cardiac frequency & $\omega = 2\pi f$ & 6.97 & rad/s & \cite{Olstad2019CiliaryDevelopment}\\
        Cilia tangential traction & $\tau$ & \num{6.5e-04} & Pa & * \\
        Pressure amplitude & $A$ & \num{1.5e-03} & Pa & * \\
        Exosome diffusion coefficient & $D_1$ & $\num{1.63e-12}$ & $\mathrm{m^2/s}$ & \textdagger\\
        STM+GFP diffusion coefficient  & $D_2$ & $\num{5.75e-11}$ & $\mathrm{m^2/s}$ & \textdagger\\
        Dendra2 diffusion coefficient  & $D_3$ & $\num{1.15e-10}$ & $\mathrm{m^2/s}$ & \textdagger\\
    \bottomrule
    \end{tabular}
    \caption{Physical parameters of the flow and transport models.
    (*) Calibrated based on experimental data~\cite{Olstad2019CiliaryDevelopment}.
    (\textdagger) Estimated based on experimental
    data~\cite{GuraSadovsky2017MeasurementExpansion, Swaminathan1997PhotobleachingDiffusion, Potma2001ReducedCells}.}
    \label{tab:parameter_table}
\end{table}

\subsection*{Computational flow modelling scenarios}
To study the impact of the motile cilia and the cardiac pulsations on the CSF flow, we consider the following computational flow scenarios, each predicting a CSF flow velocity $\uu$ and CSF pressure $p$.
\begin{itemize}
    \item A \emph{baseline} flow model representing a wild-type zebrafish including contributions both from the cilia forces and the cardiac pulsations.
    In this scenario, we apply the tangential traction $\btau$ on the cilia boundary $\Gc$ and prescribe the normal pressure $\Tilde{p}$ on the anterior/posterior boundary $\Gp$. % Model 0

    \item A \emph{cilia-only/no-cardiac} flow model including contributions from the cilia forces but no cardiac pulsations.
    Here, we apply the tangential traction $\btau$ on the cilia boundary $\Gc$, and \mer{XXX on the anterior/posterior boundary $\Gp$.} % Model I

>>>>

\section*{Methods}
\subsection*{Image-based computational geometries of zebrafish brain ventricles}
To model flow and transport in zebrafish larval brain ventricles, we generated a
3D representation of the brain ventricular walls using confocal imaging data of a zebrafish embryo
injected intraventricularly with a 70 kDa dye at 2 dpf~\cite{Olstad2019CiliaryDevelopment}. 
From the imaging data, we constructed a computational mesh of the ventricles (\Cref{fig:figure1}d).
The interior volume was meshed with fTetWild~\cite{Hu2020FastWild}. 
The \emph{standard} mesh consisted of 132 134 tetrahedral cells and had a
maximal (minimal) edge length of 14.9 $\mu$m (4.68 $\mu$m). 
At 2 dpf, the ventricular system consists mainly of 3 cavities: the telencephalic (anterior),
the di-/mesencephalic (middle) and the rhombencephalic (posterior) ventricles, which are connected by ducts. 
Parts of the ventricular surfaces were marked to be lined with
motile cilia~\cite{Olstad2019CiliaryDevelopment} (\Cref{fig:figure1}d, magenta markers).
\begin{figure}%[hbt!]
    \centering
    \includegraphics[width=\linewidth]{graphics/figure1.png}
    \caption{\lyng{Experimenting with velocity colorbars, thoughts on the
    two options? Or have two separate as it is now?}}
    \label{fig:figure1}
\end{figure}

@@

\subsection*{Computational CSF dynamics}
The beating motion of the cilia generates CSF flow with steady rotational
structures~\cite{Olstad2019CiliaryDevelopment}
within the brain ventricles (\Cref{fig:figure1}a--c).
In addition, cardiac pulsations induces pulsatile CSF flow.
We model this flow of CSF in the brain ventricles by the time-dependent,
incompressible Stokes equations which read as follows.
Find the CSF velocity $\uu = \uu(\xx, t) = (u_x(\xx, t), u_y(\xx, t), u_z(\xx, t))$
and the CSF pressure $p = p(\xx, t)$ for $\xx \in \Omega$ and time $t>0$, such that
\begin{subequations}
    \begin{align}
      \rho \pdifft{\uu} - \nabla \cdot \bsig &= \mathbf{0}
      \quad \text{in } \Omega,\\
      \nabla \cdot \uu &= 0
      \quad  \text{in } \Omega,
  \end{align}
  \label{eq:stokes_eqs}% 
\end{subequations}%
where the stress tensor is defined as $\bsig(\uu, p) = 2\mu\beps(\uu) - p\mathbf{I}$
and $\beps(\uu) = \frac{1}{2}\left(\nabla \uu + (\nabla\uu)^T\right)$ is the strain-rate tensor. 
Bold-face characters denote vectors or tensors, and $\mathbf{I}$ is the identity tensor in three dimensions.
Furthermore, $\rho=1000 \ \mathrm{kg/m^3}$ is the CSF density
and $\mu=6.97\times 10^{-4} \ \mathrm{Pa\cdot s}$ is the dynamic viscosity
of the CSF~\cite{Bloomfield1998EffectsFluid}.
We introduce the tangential traction $\bsigpar$ as the
tangential component of the traction $\hat{\bsig}=\bsig\nn$:
\begin{equation*}
    \bsigpar = P_{\nn}(\hat{\bsig}) = (\mathbf{I} - \nn\otimes\nn)\hat{\bsig},
\end{equation*}
where $\nn$ is the outer unit normal of the boundary surface $\Gamma = \partial\Omega$,
and $P_{\nn}(\rr)$ is the tangential projection of a vector $\rr$ onto $\Gamma$.
We consider three types of boundary conditions:
(i) \emph{tangential traction} (no flow normal to the boundary
$\uu \cdot \nn = 0$ with a force $\btau$ applied tangentially on the boundary $\bsigpar = \btau$),
(ii) \emph{normal pressure} (a pressure $\Tilde{p}$ applied in
the normal direction: $(\mu\nabla\uu - p\mathbf{I})\nn = \Tilde{p}(t)\nn$),
and (iii) \emph{free slip} (no flow normal to the boundary
$\uu \cdot \nn = 0$ and no tangential forces $\bsigpar = {\bf 0}$).
Unless noted otherwise, the free-slip condition is applied on the
boundary. We assume that the system starts at rest: $\uu(\xx, t=0) = \mathbf{0}$.

To study the impact of the motile cilia and the cardiac pulsations on the CSF flow,
we consider the following computational flow scenarios,
each predicting a CSF flow velocity $\uu$ and CSF pressure $p$.
\begin{itemize}
    \item A \emph{baseline} flow model representing a wild-type zebrafish including 
    contributions both from the cilia forces and the cardiac pulsations.
    In this scenario, we apply the tangential traction $\btau$ on the cilia boundary
    $\Gc$ and prescribe the normal pressure $\Tilde{p}$ on the anterior/posterior boundary $\Gp$. % Model 0

    \item A \emph{cilia-only/no-cardiac} flow model including contributions from the
    cilia forces but no cardiac pulsations.
    Here, we apply the tangential traction $\btau$ on the cilia boundary $\Gc$.
    In the absence of cardiac flow, the anterior/posterior boundary $\Gp=\emptyset$. % Model I

\subsection*{Cilia-driven flow}
On the ciliated regions of the ventricle walls $\Gc$ (\Cref{fig:figure1}d, magenta markers),
we impose a constant-in-time tangential traction $\bsigpar=\btau$ to represent the net forces
of the cilia acting on the CSF.
We set $\btau(\xx, \rr) = \tau \lambda(\xx) P_{\nn}(\rr)$, where the function $\lambda(\xx)$
and the sign of the vector $\rr=\pm(1, 0, 1)$ varies depending on the
cilia region (see~\Cref{subsec:appendixA1}).
Both $\rr$ and the value $\tau=6.5\times 10^{-4}$ Pa were chosen by 
model calibration using experimental data~\cite{Olstad2019CiliaryDevelopment}.
The data indicate speeds proximate to the ventricular walls of $27.4 \pm 5.4$ \textmu m/s and
$2.6 \pm 0.6$ \textmu m/s in the dorsal and ventral regions of the middle ventricle, respectively.
In the anterior ventricle, the speeds were estimated at $4.7 \pm 1.3$ \textmu m/s.
We calibrated $\tau$ heuristically by comparing these experimental velocity values
with the simulated, maximum velocity magnitude.

\subsection*{Flow induced by cardiac pulsations}\label{subsec:pressure_bcs}
To drive a cardiac-induced pulsatile flow inside the brain ventricles, we set $\Tilde{p}(t)=0$ and
$\Tilde{p}(t)=-A\sin{\omega t}$ on anterior and posterior parts, respectively, of the
boundary $\Gp$~(\Cref{fig:figure1}d, blue markers).
Based on a measured cardiac frequency of $f=2.22$ Hz~\cite{Olstad2019CiliaryDevelopment},
the angular cardiac frequency $\omega=2\pi f=6.97$ rad/s.
The amplitude $A=1.5\times 10^{-3}$ Pa determines the magnitude of the pulsatile flow,
and was calibrated based on experimental data~\cite{Olstad2019CiliaryDevelopment}.
Similar to $\tau$, the amplitude $A$ was chosen by comparing experimental and simulated velocities.
Particle tracking data of pulsatile motion was used to determine the contributions of the
cardiac pulsations to the velocity magnitude,
as compared to the effect of cilia motion on the flow (see~\Cref{sec:appC_particle_transport}).

@@

To computationally represent protein photoconversion in a region
$\Omega_{\mathrm{pc}} \subset \Omega$ (corresponding to ROI 1),
we prescribe a given time-dependent value to the solute concentration in this region. 
Mimicking the experimental protocol, we consider the dorsal part of the
diencephalic ventricle as the subdomain $\Omega_{\mathrm{pc}}$, and set
\begin{equation}
  c(\xx, t)=\frac{\log{(1+t/a)}}{\log{(1+T/a)}},
  \quad \xx \in \Omega_{\mathrm{pc}}, \quad t > 0.
\label{eq:photoconversion_curve}
\end{equation}
The value $a=65$ was chosen to fit the fluorescence intensity curve
observed in the physical photoconversion experiments.

\subsection*{Estimation of diffusion coefficients via the Stokes-Einstein relation}
Motivated by an interest in molecules that are significant in neural development,
we study the transport of molecules resembling the Dendra2 Fluorescent Protein (Dendra2)
used in this study, the Starmaker+Green Fluorescent Protein (STM+GFP) reported in
Jeong~\emph{et al.}~\cite{Jeong2024TheZebrafish}, and Exosomes.
To estimate the diffusion coefficients of these, when and if not
available in the literature, we use the Stokes--Einstein relation:
\begin{equation}
    D = \frac{k_B T}{6\pi \mu R},
    \label{eq:D_stokes_einstein}
\end{equation}
which is an expression for the diffusion coefficient of a spherical
molecule suspended in a viscous fluid~\cite{Einstein1905UberTeilchen}.
In~\eqref{eq:D_stokes_einstein}, $k_B = \num{1.38e-23}$ J/K is the Boltzmann constant,
$T = 310$ K is the absolute temperature, $\mu$ is the dynamic viscosity of the
solvent (in this case the CSF), and $R$ is the radius of the molecule.
For the exosomes, we calculate $D$ using~\eqref{eq:D_stokes_einstein}
with $R=200$ nm. This radius covers the slowest diffusing exosomes,
since the smaller exosomes would have diffusion coefficients similar to the other molecules.
For STM+GFP, we estimated the diffusion coefficient by extrapolating
the experimentally observed diffusion coefficient of
GFP~\cite{Swaminathan1997PhotobleachingDiffusion, Potma2001ReducedCells},
assuming that if $D$ scales with $1/R$ according to~\eqref{eq:D_stokes_einstein},
$D$ would scale with the cube root of the molecule mass~\cite{Goodhill1997DiffusionGuidance}.
The resulting diffusion coefficients used in the simulations
are $D_1$ (Exosomes), $D_2$ (STM+GFP) and $D_3$ (Dendra2)~(\Cref{tab:diff_coeff_table}).

@@

\subsection*{Computational flow modelling scenarios}
